troubleshooting

### 2024-06-17 README.Rmd
change install to main
Changed github actions to work in develop branch and polished some of the plotting functions.
README files
updating README to include new functions.

---
output:
  github_document: default
  self_contained: default
---

@@

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = FALSE
)

@@

1. Making mspms generically useful. These functions are focused on making mspms 
more generally useful to a wider audience outside of the very specific workflow
traditionally used in the O’Donoghue lab. Allows for the use of other types of
upstream proteomic data processing, different peptide libraries, etc.

@@

**Making mspms generically useful**.   
1. *prepare_peaks()*:Takes two input files from PEAKS and combines them. 
2. *prepare_pd()*: prepares exported files from proteome discoverer.  
3. *calculate_all_cleavages()*: Calculates all possible cleavages for peptide
library sequences.

@@

**Calculations/ Statistics**.   
1. *mspms_log2fc()*: Calculates the log2 fold change across experimental 
conditions relative to T0.  
2.*mspms_t_tests()*: Conducts T-tests to determine if there are any peptides 
that are significantly different across experimental conditions relative to time 
0.  
3. *log2fc_t_test()*: Combines results from mspms_log2fc() and mspms_t_tests() 
into one convenient tibble.   
4. *log2fct_condition()*: Calculates the log2 fold change and t tests for the 
specified condition at a user specified time.    
5.*log2fct_time()*: Calculates the log2 fold change and t tests between the 
specified times within a user specified time.   
6. *mspms_anova()*: Conducts ANOVA tests to determine if there are any peptides 
that are significantly different across all time points.   
7. *count_cleavages_per_pos():* Counts the number of cleavages at each position 
of the library. Useful for determining endoprotease vs exoprotease activity.

**Data Visualization**.   
1. *plot_heatmap()*: Conducts hierarchical clustering analysis and plots an
interactive heatmap to visualize overall patterns in the data.    
2. *plot_pca()*: PCA analysis to visualize the data in a 2D space.  
3. *plot_time_course()*: Plot peptides over time by condition.  
4. *plot_cleavages_per_pos()*: Plot the number of cleavages at each position 
in the library.Useful for determining endoprotease vs exoprotease activity.

**Icelogo**.

1. *calc_AA_count_of_motif()*: calculates the count of each amino acid at each
position in a vector of motifs.  
2. *calc_AA_prop_of_motif()*: calculates the proportion of each amino acid at 
each position from the output of calc_AA_count_of_motif().  
3. *calc_AA_motif_zscore()*: calculates z scores for each amino acid at each 
position in a vector of motifs.  
4. *calc_AA_percent_difference()*: calculates the percentage difference of 
5. *prepare_sig_p_dif()*: prepares the data for plotting percent difference 
icelogos. Removes AAs at positions that are not significantly different.  
6. *prepare_fc()*: prepares the data for plotting fold change icelogos.
Removes AAs at positions that are not significantly different.  
7. *plot_pd_icelogo()*: plots a percent difference icelogo.  
8. *plot_fc_icelogo()*: plots a fold change ice logo.  
9. *plot_icelogo()*:  wraps plot_pd_icelogo() and plot_fc_icelogo() into 
one function.  
10. *plot_all_icelogos()*: convenience function to plot all icelogos relative to 
time 0 from an experiment.    
**Reports**. 
1. *generate_report()*: Generates a simple analysis of the mspms data that 
should be generically applicable to all mspms experiments.

## Making Generically Usefull.

@@

All analysis performed by this package is downstream of the data generated by 
the PEAKS software. The first step is to combine the two files generated by
PEAKS.

The files coming from peaks should be generated according to the instructions 
found [here](www/PeaksDataAnalysis_howto.pdf).

@@

```{r}
all_peptide_sequences <- mspms::calculate_all_cleavages(mspms::peptide_library$library_real_sequence,
  n_AA_after_cleavage = 4
)
head(all_peptide_sequences)
```

@@

```{r}
all_peptide_sequences <- mspms::calculate_all_cleavages(mspms::peptide_library$library_real_sequence,
  n_AA_after_cleavage = 5
)
head(all_peptide_sequences)
```

@@

We could also look at the fold change instead of the percent difference

```{r}
mspms::plot_icelogo(cleavage_seqs, background_universe, type = "fold change")
```

@@

This function includes cleavages that were significant at any time point once in 
the experimental set of the ICELOGO.

```{r}
mspms::plot_all_icelogos(mspms_data)
```

@@

![](plot_heatmap_output.png)

>>>>

---
output: github_document

---

@@

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

@@

1. Pre-processing data. These functions are focused on making mspms 
more generally useful to a wider audience outside of the very specific workflow
traditionally used in the O’Donoghue lab. Allows for the use of other types of
upstream proteomic data processing, different peptide libraries, etc.

@@

**Preprocessing data**.   
1. *prepare_peaks()*:Takes two exported files from PEAKS software, 
combines them, and prepares for processing by the mspms R package. 
2. *prepare_pd()*: prepares exported files from proteome discoverer.  
3. *calculate_all_cleavages()*: Calculates all possible cleavages for peptide
library sequences.

@@

**Calculations/ Statistics**.   
1. *mspms_log2fc()*: Calculates the log2 fold change across experimental 
conditions relative to T0.  
2.*mspms_t_tests()*: Conducts T-tests to determine if there are any peptides 
that are significantly different across experimental conditions relative to time 
0.  
3. *log2fc_t_test()*: Combines results from mspms_log2fc() and mspms_t_tests() 
into one convenient tibble.   
4. *log2fct_condition()*: Calculates the log2 fold change and t tests for the 
specified condition at a user specified time.    
5.*log2fct_time()*: Calculates the log2 fold change and t tests between the 
specified times within a user specified time.   
6. *mspms_anova()*: Conducts ANOVA tests to determine if there are any peptides 
that are significantly different across all time points.   
7. *count_cleavages_per_pos():* Counts the number of cleavages at each position 
of the library. Useful for determining endoprotease vs exoprotease activity.
8. *count_cleavages_per_pos()*: Counts the number of cleavages at the position
of the initial peptide detected peptides were cleaved from.

**Icelogo calculations**

1. *calc_AA_count_of_motif()*: calculates the count of each amino acid at each
position in a vector of motifs.  
2. *calc_AA_prop_of_motif()*: calculates the proportion of each amino acid at 
each position from the output of calc_AA_count_of_motif().  
3. *calc_AA_motif_zscore()*: calculates z scores for each amino acid at each 
position in a vector of motifs.  
4. *calc_sig_zscore()*: given a matrix of zscores, determines which are significant given a user defined p value. 
5. *calc_AA_percent_difference()*: calculates the percentage difference of 
6. *calc_AA_fold_change()*: calculates the fold change of a matrix containing 
the proportions of amino acids at each position relative to a background matrix. 
7. *prepare_sig_p_dif()*: prepares the data for plotting percent difference 
icelogos. Removes AAs at positions that are not significantly different.  
8. *prepare_pd()*:prepares a matrix with significant amino acid percent 
differences by position for plotting.
9. *prepare_fc()*: prepares the data for plotting fold change icelogos.
10. *prepare_icelogo_data()*: uses all the functions described above to 
prepare the data for plotting icelogos. 
11. *extract_re()*: extracts regular expressions specifying amino acids above a
defined threshold cutoff for each position of a cleavage motif from an icelogo 
matrix.

**Data Visualization**.   
1. *plot_heatmap()*: Conducts hierarchical clustering analysis and plots an
interactive heatmap to visualize overall patterns in the data.    
2. *plot_pca()*: PCA analysis to visualize the data in a 2D space.  
3. *plot_time_course()*: Plot peptides over time by condition.  
4. *plot_cleavages_per_pos()*: Plot the number of cleavages at each position 
in the library.Useful for determining endoprotease vs exoprotease activity.
5.*plot_icelogo()*: Plot an icelogo given an icelogo matrix. 
6.*plot_all_icelogos()*: generates icelogos representing enriched sequences 
relative to time 0 for each conditions. This function compares cleavage 
sequences corresponding to peptides that are significant 
(p.adj < 0.05 and log2 fold change > 3) to a background of all possible
sequences present in the initial peptide library.

**Reports**. 
1. *generate_report()*: Generates a simple analysis of the mspms data that 
should be generically applicable to all mspms experiments.

## Data Pre-processing

@@

Analysis downstream of data exported from PEAKS software is supported.

exported files are generated in the O'donoghue lab as follows:

1. Create a New Project in PEAKS.  
2. Select data to add to PEAKS project.  
3. Add data, rename samples, set correct parameters. 
  • Highlight all the added data.  
  • Select Create new sample for each file. 
  • Select appropriate instrument and fragmentation method. 
  •  Set Enzyme to “None”. 
  • Rename samples according to enzymes, time points, and replicates. 
      • Can also ensure appropriate data is selected for these   
  • Select “Data Refinement” to continue. 
4. Data Refinement. 
  • On the top right, select MSP-MS as the predefined parameters. 
  •  Parameters shown are what should be used for MSP-MS data analysis. 
  •  Select “Identification” to continue. 
5. Peptide Identification. 
  • On the top right, select MSP-MS as the predefined parameters. 
  • Parameters shown are what should be used for MSP-MS data analysis. 
  • Identification -> Select Database -> TPD_237. 
  • Identification -> no PTMs. 
  • Can remove the PTMs by highlighting them and selecting Remove  
6. Label Free Quantification. 
  • Make sure Label Free is selected. 
  • Group samples -> add quadruplicates of samples to new group.

@@

```{r}
all_peptide_sequences <- mspms::calculate_all_cleavages(
  mspms::peptide_library$library_real_sequence,
  n_AA_after_cleavage = 4
  )
head(all_peptide_sequences)
```

@@

```{r}
all_peptide_sequences <- mspms::calculate_all_cleavages(
  mspms::peptide_library$library_real_sequence,
  n_AA_after_cleavage = 5
)
head(all_peptide_sequences)
```

@@

We could also look at the fold change instead of the percent difference.
```{r}
mspms::plot_icelogo(cleavage_seqs, background_universe, type = "fold_change")
```

@@

This function includes cleavages that were significant at any time point once in 
the experimental set of the ICELOGO. Below we show this applied to create
percent difference and fold change icelogos.

```{r}
plot_all_icelogos(mspms_data)

plot_all_icelogos(mspms_data, type = "fold_change")
```

### Extracting regular expressions from icelogos

It might be useful for the user to extract the information contained in the 
Icelogo and use it to filter their data for peptides that are predicted 
to be well cleaved, or not cleaved according to the icelogo plot.

For example, we can get the amino acids that are enriched at each position of 
the icelogo using the code below.

# Getting data for Icelog 
cleavage_seqs <- time_t %>%
  dplyr::filter(log2fc > 3, p.adj <= 0.05) %>%
  dplyr::pull(cleavage_seq)

# Plotting icelogo
mspms::plot_icelogo(cleavage_seqs, background_universe)

# corresponding icelogo data to the plot
icelogo_matrix = mspms::prepare_icelogo_data(cleavage_seqs, background_universe)

# Extracting regular expressions
RE_pos = mspms::extract_re(icelogo_matrix)

RE_pos
```

We can also do this for amino acids that are negatively enriched.

```{r}
RE_neg = mspms::extract_re(icelogo_matrix,type = "negative")

RE_neg
```

Once we have these regular expressions, we can use it to filter our data as
follows:

```{r}
pos_enriched = mspms_data %>% 
  dplyr::filter(grepl(RE_pos, cleavage_seq))
```

In this case, none of the peptides in the original data set match all the rules
in our library (peptide library does not contain all possible combinations, so
none of the peptides was a match to the combination of rules at each position)

We could then filter to only include peptides that are greater than a certain
threshold percent difference. We can then visualize these results easily with 
the plot_time_course function described further below. Here we will only 
consider AAs with a percent difference greater than 10 at each position.

```{r}
# Extracting regular expressions 
RE_pos = mspms::extract_re(icelogo_matrix,threshold = 10)

RE_pos

# Filtering data for peptides that have cleavage sequences matching regular 
# expressions
f = mspms_data %>% 
  dplyr::filter(grepl(RE_pos, cleavage_seq))

#visualizing these peptides that match over time.
plot_time_course(f)
```

Here we see that these peptides were cleaved over time in DMSO, as would be 
expected based on the icelogo plot.

@@

![](README_files/plot_heatmap_output.png)
